Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of mixed binary observations

### 2016-06-10 Paper/paper.tex
updated figs

% #################
\begin{figure*}
\centering 
\includegraphics[width=1.8\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI90_0_Lambda_SNR30_70_linear}
\caption{We show here the ratio of systematic and statistical
measurement uncertainties for the binary chirp mass over the NSBH parameter 
space. Each panel shows the same as a function of BH mass and spin. Across
each row, we can see the effect of increasing the signal strength (i.e. SNR)
with the NS's deformability fixed. Down each column, we can see 
the effect of the increasing NS's tidal deformability at fixed SNR. We also
show dashed contours for where the ratio equals $10\%$ (labelled ``-1'')
or $100\%$ (labelled ``0'').
% 
Generally, these two sources of uncertainty are different and for BBHs, the
statistical errors dominate systematic ones~\cite{Kumar:2016dhh}. We find
that for BHNS binaries, its not much different until we get to high SNRs
$\rho\gtrsim 70$ for which if the BH is spinning rapidly enough, the 
measurement uncertainty for $\mchirp$ could become dominated by the exclusion
of tidal effects in filtering templates.
}
\label{fig:TN_chirpMassBias_vs_Lambda_SNR}
\end{figure*}
%
\begin{figure*}[!t]
\centering
\includegraphics[width=1.8\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI90_0_Lambda_SNR30_70_linear002}
\caption{This figure is similar to Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}
with the difference that the quantity in question here is the symmetric 
mass-ratio $\eta$. We find that for fairly loud GW signals, with $\rho\simeq 50$,
not including the effects of NS's tidal deformation on GW emission can become the dominant
source of error for astrophysical searches with Advanced LIGO. However,
for quieter signals with $\rho\leq 30$, it will have minimal effects on the measurement
of $\eta$. We remind the reader that the SNRs here are all single detector values.
}
\label{fig:TN_EtaBias_vs_Lambda_SNR}
\end{figure*}
% 
% 
\begin{figure*}
\centering
\includegraphics[width=1.95\columnwidth]{plots/TNChiBHBiasesOverCIWidths_CI90_0_Lambda_SNR_linear}
% \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI68_3.pdf}
% \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI90_0.pdf}
\caption{This figure shows the ratio of the systematic and statistical
measurement errors for BH spins, with other attributed identical to 
Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}, and~\ref{fig:TN_EtaBias_vs_Lambda_SNR}.
Similar to the case of mass parameters, we find that below
$\rho\approx 30$, ignoring tidal effects in templates introduces minor systematic effects,
which remain subdominant to the statistical measurement uncertainties.
}
\label{fig:TN_BHspinBias_vs_Lambda_SNR}
\end{figure*}
% 
% 
Past and future GW search and source characterization efforts with Advanced LIGO
have used (or plan to) BH-BH waveform templates for NS-BH mergers, ignoring
NS tidal effects on the emitted GWs. In this section we present a fully
Bayesian analysis of the effects of this assumption on the recovery of NSBH
masses and spins from their GW signals.

@@

\includegraphics[trim={3cm 0 0 0},width=1.9\columnwidth]{plots/TTLambdaRawCIWidths90_0_Lambda_SNR}

@@

In the previous section we showed that we begin to care for the effects of NS's
deformation due to its companion hole's tidal field, if (a) the BH is sufficiently
small, (b) BH spin is aligned and $\gtrsim +0.4$, and (c) the NS is
not very compact, with $\lambdans\gtrsim 1000$. The first two conditions
enhance tidal effects on emitted GWs by increasing the number of orbits at small
separation, where the deviation of NSBH and BHBH dynamics is the largest. The third,
i.e. (c), simply makes the NS more easily deformable. All of them reduce the onset 
frequency for NS's disruption close to merger. Therefore, systems that fulfil the 
above conditions have considerably (or, at least, the most) different GW morphology
when compared to equivalent BHBH coalescences. Ultimately, the extent of 
tidal modulations of emitted GWs depends on the equation of state of neutron star
matter, which becomes an indirect measurable in GW astrophysics.
% 
However, if we do indeed receive GWs from such a binary source, what information can we
possibly extract from it about the internal structure of neutron stars?
In this section we consider this question in perspective of measuring the tidal 
deformability parameter of neutron stars purely from gravitational waves they emits 
while orbiting spinning black holes.
% 
We find that when the BH is sufficiently spinning in the prograde sense,
with $\chibh\gtrsim +0.7$, even at a relatively possible signal strengths ($\rho=20$),
we can begin to constraint $\lambdans$ at the level of $\pm 75\%$ errors with a $90\%$
confidence level. If the signal is stronger, say $\rho=30$, we can contraint
$\lambdans$ better with only $\pm 50\%$ errors, for binaries with large BH spins.
This trend continues as we vary the SNR from $\rho=30-50$ upwards. In other words,
with a single but moderately loud NSBH signal observation, Advanced LIGO will begin
to put constraints on NS matter properties. These constraints can consequently
be used to assess the likelihood of different equations of state for nuclear
matter, and possibly narrow the range they span.

  `

We use the same set of systems as in the previous section, with $q=\{2,3,4,5\}$,
$\chibh=\{-0.5,0,+0.5,+0.75\}$, $\lambdans=\{500, 800, 1000, 1500, 2000\}$ and SNRs
$\rho=\{20,30,50,70\}$. For each unique combination of these parameters, we obtain a
signal waveform using the tidal model of Lackey et al~\cite{Lackey:2013axa}.
We use the same waveform model as input to our Bayesian parameter estimation
method in order to obtain estimates for different binary parameters alongwith 
their full probability distributions, including for the combined NS tidal
deformability parameter $\lambdans$.

The precision with which $\lambdans$ can be measured is given by the width of 
its $90\%$ confidence interval, as a fraction of the true value. 
In Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR} we show the same as a function
of binary parameters and signal strength. Each panel shows the percent statistical
uncertainty $\dlambda$ in the measurement of $\lambdans$ as a function of BH 
masses and spins, with NS parameters and signal strength being held fixed. 
In each row, the signal strength is increased from
left to right with $\rho = 20\rightarrow 70$. In each column, the deformability
of the NS is increased from top to bottom, with $\lambdans=500\rightarrow 2000$.
% 
We find that at SNRs around $\rho=30$, we begin to constrain $\lambdans$ 
meaningfully with $\dlambda\approx 100-150\%$. Amongst other parameters, 
the BH spin and mass play a big role. A smaller BH with a larger spin always
allows for a more precise measurement on $\lambdans$. Visually we can see that
in each panel the bottom right corner, which corresponds to low-mass BHs
with large spins, is the region of smallest measurement errors on $\lambdans$.
The actual deformability of the NS also plays an important role on its own
measurability. For e.g., when $\lambdans\leq 1000$, it is fairly difficult
to meaningfully constrain $\lambdans$ without getting remarkably lucky and
finding such a binary with $\rho\gtrsim 50$.

>>>>

% #################
\begin{figure*}
\centering 
\includegraphics[width=1.8\columnwidth]{plots/TNMchirpBiasesOverCIWidths_CI90_0_Lambda_SNR30_70_linear}
\caption{We show here the ratio of systematic and statistical
measurement uncertainties for the binary chirp mass over the NSBH parameter 
space. Each panel shows the same as a function of BH mass and spin. Across
each row, we can see the effect of increasing the signal strength (i.e. SNR)
with the NS's deformability fixed. Down each column, we can see 
the effect of the increasing NS's tidal deformability at fixed SNR. We also
show dashed contours for where the ratio equals $10\%$ (labelled ``-1'')
or $100\%$ (labelled ``0'').
% 
Generally, these two sources of uncertainty are different and for BBHs, the
statistical errors dominate systematic ones~\cite{Kumar:2016dhh}. We find
that for BHNS binaries, its not much different until we get to high SNRs
$\rho\gtrsim 70$ for which if the BH is spinning rapidly enough, the 
measurement uncertainty for $\mchirp$ could become dominated by the exclusion
of tidal effects in filtering templates.
}
\label{fig:TN_chirpMassBias_vs_Lambda_SNR}
\end{figure*}
%
\begin{figure*}[!t]
\centering
\includegraphics[width=2\columnwidth]{plots/TNEtaBiasesOverCIWidths_CI90_0_Lambda_SNR20_70_linear}
\caption{This figure is similar to Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}
with the difference that the quantity in question here is the symmetric 
mass-ratio $\eta$. We find that for fairly loud GW signals, with $\rho\simeq 50$,
not including the effects of NS's tidal deformation on GW emission can become the dominant
source of error for astrophysical searches with Advanced LIGO. However,
for quieter signals with $\rho\leq 30$, it will have minimal effects on the measurement
of $\eta$. We remind the reader that the SNRs here are all single detector values.
}
\label{fig:TN_EtaBias_vs_Lambda_SNR}
\end{figure*}
% 
% 
\begin{figure*}
\centering
\includegraphics[width=1.95\columnwidth]{plots/TNChiBHBiasesOverCIWidths_CI90_0_Lambda_SNR_linear}
% \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI68_3.pdf}
% \includegraphics[width=\columnwidth]{plots/TNChiBHBiasesOverCIWidthsVsSNR_All_CI90_0.pdf}
\caption{This figure shows the ratio of the systematic and statistical
measurement errors for BH spins, with other attributed identical to 
Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}, and~\ref{fig:TN_EtaBias_vs_Lambda_SNR}.
Similar to the case of mass parameters, we find that below
$\rho\approx 30$, ignoring tidal effects in templates introduces minor systematic effects,
which remain subdominant to the statistical measurement uncertainties.
}
\label{fig:TN_BHspinBias_vs_Lambda_SNR}
\end{figure*}
% 
% 
Past and future GW search and source characterization efforts with Advanced LIGO
have used (or plan to) BH-BH waveform templates for NS-BH mergers, ignoring
NS tidal effects on the emitted GWs. In this section we present a fully
Bayesian analysis of the effects of this assumption on the recovery of NSBH
masses and spins from their GW signals.

@@

\includegraphics[trim={2cm 0 0 0},width=2.2\columnwidth]{plots/TTLambdaRawCIWidths90_0_Lambda_SNR}

@@

In the previous section we showed that we begin to care for the effects of NS's
deformation due to its companion hole's tidal field, if (a) the BH is sufficiently
small, (b) BH spin is aligned and $\gtrsim +0.4$, and (c) the NS is
not very compact, with $\lambdans\gtrsim 1000$. The first two conditions
enhance tidal effects on emitted GWs by increasing the number of orbits at small
separation, where the deviation of NSBH and BHBH dynamics is the largest. The third,
i.e. (c), simply makes the NS more easily deformable. All of them reduce the onset 
frequency for NS's disruption close to merger. Therefore, systems that fulfill the 
above conditions have considerably (or, at least, the most) different GW morphology
when compared to equivalent BHBH coalescences. Ultimately, the extent of 
tidal modulations of emitted GWs depends on the equation of state of neutron star
matter, which becomes an indirect measurable in GW astrophysics.
% 
However, if we do indeed receive GWs from such a binary source, what information can we
possibly extract from it about the internal structure of neutron stars?
In this section we consider this question in perspective of measuring the tidal 
deformability parameter of neutron stars purely from gravitational waves they emits 
while orbiting spinning black holes.
% 
\red{Move to discussion}
\prayush{%
We find that when the BH is sufficiently spinning in the prograde sense,
with $\chibh\gtrsim +0.7$, even at a relatively possible signal strengths ($\rho=20$),
we can begin to constraint $\lambdans$ at the level of $\pm 75\%$ errors with a $90\%$
confidence level. If the signal is stronger, say $\rho=30$, we can contraint
$\lambdans$ better with only $\pm 50\%$ errors, for binaries with large BH spins.
This trend continues as we vary the SNR from $\rho=30-50$ upwards. In other words,
with a single but moderately loud NSBH signal observation, Advanced LIGO will begin
to put constraints on NS matter properties. These constraints can consequently
be used to assess the likelihood of different equations of state for nuclear
matter, and possibly narrow the range they span.
}
  `
\red{Is this the LEA+ mode? If yes, say so}
\prayush{%
We use the same set of systems as in the previous section, with $q=\{2,3,4,5\}$,
$\chibh=\{-0.5,0,+0.5,+0.75\}$, $\lambdans=\{500, 800, 1000, 1500, 2000\}$ and SNRs
$\rho=\{20,30,50,70\}$. For each unique combination of these parameters, we obtain a
signal waveform using the tidal model of Lackey et al~\cite{Lackey:2013axa}.
We use the same waveform model as input to our Bayesian parameter estimation
method in order to obtain estimates for different binary parameters along with 
their full probability distributions, including for the NS tidal
deformability parameter $\lambdans$.
}

\red{Present posterior distributions of selected configurations first..}

\red{%
\begin{enumerate}
 \item State the prior on $\lambdans$ because $\lambdans$ weakly deforms waveforms,
I would expect that measurement uncertainty in $\lambdans$ is approx. independent
of $\lambdans$, except for running into the $\lambdans\geq 0$ prior.
 \item Is $\lambdans$ measurement degenerate with other parameters? Either cite 
 known studies [if any], or provide 2-d posterior plots.
\end{enumerate}
}
The precision with which $\lambdans$ can be measured is given by the width of 
its $90\%$ confidence interval, as a fraction of the true value. 
In Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR} we show the same as a function
of binary parameters and signal strength. Each panel shows the percent statistical
uncertainty $\dlambda$ in the measurement of $\lambdans$ as a function of BH 
masses and spins, with NS parameters and signal strength being held fixed. 
In each row, the signal strength is increased from
left to right with $\rho = 20\rightarrow 70$. In each column, the deformability
of the NS is increased from top to bottom, with $\lambdans=500\rightarrow 2000$.
% 
We find that at SNRs around $\rho=30$, we begin to constrain $\lambdans$ 
meaningfully with $\dlambda\approx 100-150\%$. Amongst other parameters, 
the BH spin and mass play a big role. A smaller BH with a larger spin always
allows for a more precise measurement on $\lambdans$. Visually we can see that
in each panel the bottom right corner, which corresponds to low-mass BHs
with large spins, is the region of smallest measurement errors on $\lambdans$.
The actual deformability of the NS also plays an important role on its own
measurability. For e.g., when $\lambdans\leq 1000$, it is fairly difficult
to meaningfully constrain $\lambdans$ without getting remarkably lucky and
finding such a binary with $\rho\gtrsim 50$.
